Learning from Data Through Models

### 2023-07-11 paper/paper.Rmd
writing learning results

---
title: "Learning from Data Through Models"
date: '`r format(Sys.time(), "%B %d, %Y")`'
author: "Alberto Bisin, Guillaume Frechette and Jimena Galindo"
abstract: "TBW"
output: 
    pdf_document:
        citation_package: natbib
bibliography: bibliography.bib
fontsize: 12pt
geometry:
- margin=2.5cm
hyperref:
- linestretch=2.0
documentclass: article
---

```{r dependencies, include=FALSE}
library(data.table)
library(here)
renv::snapshot()

@@

## Optimal Rules
The most accurate prediction rule for this setting will be one that predicts the state to be **B** whenever, 
given the observables, **B** is more likely. And it predicts **R** whenever **R** is more likely 
given the observables. Because we abstract away from the procedure through which the rule is updated, 
we will make predictions under the assumption that the DM has learned the parameters perfectly. That is, 
they have seen enough information that, under an updating procedure that converges to the truth---for example, 
correctly specified regression model or Bayes rule---they would have converged to the true parameters already.

@@

## Polarization
To study polarization, we take the definition of @Haghtalab2021: two DMs are said to be 
polarized if, when observing the same realization of observables, they predict the state to be different. That is, 
they both observe the same draw of $x=(x_1, ..., x_N)$ but one DM predicts the state to be **B** and the other 
predicts the state to be **R**.

@@

#### Example
_The optimal rule for the complex model coincides with the truth, it predicts **B** if $x_2+1\geq x_1$. 
In addition, there are two simple models: one that considers only $x_1$---call it $M_1$---and one that considers 
only $x_2$---call it $M_2$. The optimal rules for $M_1$ and $M_2$ are threshold rules that predict the state is **B** 
if the value of $x_i$ is below (above) the threshold $t_i := \mu_{-i} + k$ and the predicted state is **R** otherwise.
These prediction rules are illustrated in . As the figure makes clear, both rules agree on the predicted state for 
some values of $(x_1, x_2)$ and disagree for other values. Theregions in which the rules make different predictions
are the disagreement zones._

@@

# Experimental Design
```{r read-data-all}
data <- fread(here("data", "clean", "all.csv"))
```

@@

Part 2 started immediatly after part 1. In this part each subject was randomly allowed to use only a certain number of variables
in order to make their prediction. Instead of showing them the realizations of all 5 variables, they were told that they can 
disclose up to $K$ variables, where $K$ was drawn at random from the set $\{1, 2, ..., 5\}$ at every round and it was drawn 
independently across subjects. We take the variables that they chose to reveal as the model that each subject using for predicting 
the state and use that model to detrmine whether the theory would predict the pair to be polarized or not. In addition, we use 
the random assignment of $K$ to investigate whether the subjects are using siumple or complex models. If allowing subjects to 
disclose an additional variable improves their prediction accuracy, then it must be that they are in fact using the additional
information for prediciton. If, on the other hand, they do not perform better when they have access to more information, it must be
that although they are disclosing an additional variable, they do not correctly incorporate the infomation into their prediciton 
rule. Thus, if we observe that the performance of our subjects does not improve when they are allowed to disclose $K+1$ variables, 
relative to their performance when they were allowed to disclose only $K$ variables, their model of the world must be of size $K$
or lower. This feature design allows us to test prediciton 1 and 1B and determine the number of variables that subjects are using.

# Data

@@

>>>>

---
title: "Learning from Data Through Models"
date: '`r format(Sys.time(), "%B %d, %Y")`'
author: "Alberto Bisin, Guillaume Frechette and Jimena Galindo"
abstract: "TBW"
output: 
    pdf_document:
        citation_package: natbib
        fig_width: 7
        fig_height: 6
        fig_caption: true
        number_sections: false
        keep_tex: true
        extra_dependencies:
            footmisc: ["bottom"] # footnote management
            setspace: ["doublespacing"] # spacing of the paper
            caption: ["normal"]
            dsfont: null # indicator function 1
            booktabs: null
            makecell: null
            hyperref: null

bibliography: bibliography.bib
fontsize: 12pt
geometry:
- margin=2.5cm
hyperref:
- linestretch=2.0
documentclass: article
---

```{r dependencies, include=FALSE}
library(data.table)
library(here)
renv::snapshot()
library("languageserver")
library(reticulate)
```

@@

### Optimal Rules
The most accurate prediction rule for this setting will be one that predicts the state to be **B** whenever, 
given the observables, **B** is more likely. And it predicts **R** whenever **R** is more likely 
given the observables. Because we abstract away from the procedure through which the rule is updated, 
we will make predictions under the assumption that the DM has learned the parameters perfectly. That is, 
they have seen enough information that, under an updating procedure that converges to the truth---for example, 
correctly specified regression model or Bayes rule---they would have converged to the true parameters already.

@@

### Polarization
To study polarization, we take the definition of @Haghtalab2021: two DMs are said to be 
polarized if, when observing the same realization of observables, they predict the state to be different. That is, 
they both observe the same draw of $x=(x_1, ..., x_N)$ but one DM predicts the state to be **B** and the other 
predicts the state to be **R**.

@@

##### Example
_The optimal rule for the complex model coincides with the truth, it predicts **B** if $x_2+1\geq x_1$. 
In addition, there are two simple models: one that considers only $x_1$---call it $M_1$---and one that considers 
only $x_2$---call it $M_2$. The optimal rules for $M_1$ and $M_2$ are threshold rules that predict the state is **B** 
if the value of $x_i$ is below (above) the threshold $t_i := \mu_{-i} + k$ and the predicted state is **R** otherwise.
These prediction rules are illustrated in . As the figure makes clear, both rules agree on the predicted state for 
some values of $(x_1, x_2)$ and disagree for other values. Theregions in which the rules make different predictions
are the disagreement zones._

@@

# Experimental Design
```{r read-data-all}
data <- fread(here("data", "clean", "all.csv"))
part1 <- fread(here("data", "clean", "part1.csv"))
part2 <- fread(here("data", "clean", "part2.csv"))

@@

Part 2 started immediatly after part 1 and has 70 rounds. In this part each subject was randomly allowed to use only a certain 
number of variables in order to make their prediction. Instead of showing them the realizations of all 5 variables, they were 
told that they can disclose up to $K$ variables, where $K$ was drawn at random from the set $\{1, 2, ..., 5\}$. K was redrawn
independently every round and across subjects. We use the random assignment of $K$ to investigate whether the subjects are using 
siumple or complex models. If allowing subjects to disclose an additional variable improves their prediction accuracy, 
then it must be that they are in fact using the additional information for prediciton. If, on the other hand, they do not perform 
better when they have access to more information, it must be that although they are disclosing an additional variable, 
they do not correctly incorporate the infomation into their prediciton rule. Thus, if we observe that the performance of our 
subjects does not improve when they are allowed to disclose $K+1$ variables, 
relative to their performance when they were allowed to disclose only $K$ variables, their model of the world must be of size $K$
or lower. This feature design allows us to test prediciton 1 and 1B and determine the number of variables that subjects are using.

In order to explore the polarization prediciton, we use the fact that subjects are in fixed pairs throughout the experiment.
Both subjects in a pair observe the same realization of all the variables in each round. Therefore, we can determine whether
they are polarized or not by looking at whether they make the same prediction or not. We can then determine whether the theory 
predicts that they should be polarized or not by looking at the polarization zone corresponding to the models that they are using.
If the draw of the variables falls into the polarization zone, then the theory predicts that they should be polarized. The models
that we use for this prediction are the ones that include the variables that they chose to reveal in that round. It could be that
although they are revealing certain variables, they are not using them for prediction, however, we do not have a good way to 
determine this. Instead, we use the variables that they reveal as a proxy for the variables that they are using for prediction and 
 determine the polarization zones for those models.

@@

In this section we present the results from the experiment. We start by exploring the general learning patterns of the subjects
as well as the model choices. We then look at the effect of allowing subjects to disclose an additional variable and lastly we
look at the polarization results.

### Learning
In the first part of the experiment, subjects were asked to predict the state given the realization of all the variables.
Having access to all the information they could predict the state perfectly if they managed to learn the parameters of the
hyperplane that determines the state. Figure \ref{learningAll} shows the share of correct predictions across rounds in part 1. 
We see that subjects are able to learn how to predict the state at a rate that is higher than random. They initially predict
the state corectly about 50% of the time, which is consistent with random guessing. The share of correct guesses increases to 
`r 100*round(part1[subsession.round_number>15, mean(player.correct)], digits=2)`% by the end of part 1. And this last number is 
significantly higher than 50% 
(p-value = `r round(t.test(part1[subsession.round_number>10, player.correct], mu=0.5)$p.value, digits=4)`). 
They continue to learn even after the first 20 rounds when we look at the rounds in which subjects had access to all the information.
By the end of the experiment the rate of correct predictions is 
`r 100*round(part2[subsession.round_number>30, mean(player.correct)], digits=2)`% which is still significantly higher than 50%
(p-value = `r round(t.test(part2[subsession.round_number>30, player.correct], mu=0.5)$p.value, digits=4)`).

```{r learningAll, out.width='50%', fig.align='center', fig.cap='\\label{learningAll} Share of correct predictions across rounds'}
knitr::include_graphics(here("computed_objects", "figures", "learning_all.png"))
```
